ExEl: Building an Elaborator Using Extensible Constraints

### 2024-03-04 paper/main.tex
Picking some nits in the abstract

  There has been a lot of work to find the right theoretical foundations for the core type theory underlying proof assistants and dependently-typed programming languages like Coq, Agda, Lean or Idris.
  However, their implementations have been studied to a smaller extent.
  Together with the additions of features that aren't reflected in theoretical studies, this means that developers have to make a lot of adaptations in the implementations, making the codebases that are hard to maintain.
  In this work we consider some common problems in the implementations of the type-checkers and propose a new architecture that allows for a more modular development and containment of features to one module each.
  We focus focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax with a lot of features into a small well-behaved core theory.
  Our core idea is to use an open datatype for constraints and a plugin system for solvers.
  We showcase the design on a proof-of-concept implementation.

>>>>

  Proof assistants and dependently typed languages such as Coq, Agda, Lean, and Idris can be used to ascertain the correctness of software with mathematical precision.
  While much research has been done on their theoretical foundations, their actual implementations have been studied to a much lesser extent.
  As a result, features that are not considered part of the theoretical foundations - such as implicit arguments and type classes - have their own bespoke implementation for each language, making for code bases that are hard to understand and maintain.
  To address some of the common problems in the implementations of dependently typed languages, we present a modular architecture for implementing the transformation from user-friendly surface syntax into a small and well-behaved core language, also known as elaboration.
  Our architecture is made modular through the use of an open type of constraints and a plugin system for solvers that work on these constraints, which means that each new feature is contained in its own module.
  We showcase our design with a proof-of-concept elaborator for a language with dependent types, implicit arguments, higher-order unification, implicit coercions, and instance arguments.
